---
title: "2016 Election"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Margin of Victory

Let's take a look at the 2016 electoral map, divided by county.
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
library(reshape2)
library(tigris)
library(ggplot2)
library(leaflet)
library(maps)
library(scales)
library(rgdal)
library(rgeos)
library(sqldf)

####General Notes####
# Use the section-picker at the bottom of the script window to jump between code sections
# Our analysis only considered the lower 48 states and the District of Columbia (excluded Alaska, Hawaii and overseas territories)
# The tigris data requires the below global option if you are running Windows 10
options(tigris_use_cache = FALSE)

####2016 General Election Data####
setwd("~/Second Year/DS 4559 - Data Science/Final Project 2/election2016/Data")

#Read the 2016 general election data + FIPS code database
gen16 <- read.csv("pres16results.csv", header = TRUE, stringsAsFactors = FALSE)
fips.labels <- read.csv("fips_database.csv", header = FALSE, stringsAsFactors = FALSE, colClasses = "character")

#Cleaning general election data
gen16 <- gen16[which(gen16$cand == "Donald Trump" | gen16$cand == "Hillary Clinton"),]
gen16 <- gen16[-c(1,2),]

sapply(gen16,class)

#This county wasn't labeled in the data, so we manually assigned it the county name
for (i in seq(nrow(gen16))){
  if (gen16$fips[i] == "46102") gen16$county[i] <- "Oglala Lakota County"
}

#Creating a separate table with just the state-wide results
gen16.states <- gen16[is.na(gen16$county),]
gen16.states <- gen16.states[-c(103,104),] #removing state FIPS for Alaska
gen16.states$county <- NULL
gen16.states$fips <- NULL

gen16 <- gen16[-is.na(gen16$county)]

#Removing the statewide results from the county data
for (i in seq(nrow(gen16))) {
  if (nchar(gen16$fips[i])<=2) gen16$fips[i] <- NA
}

gen16 <- gen16[-which(is.na(gen16$fips)),]
gen16$pct_report <- NULL
gen16.2 <- gen16

gen16.2$votes <- NULL
gen16.2$lead <- NULL

#Used the reshape2 package to convert the dataframe from long to wide format
long16 <- dcast(gen16.2, fips + st + total_votes ~ cand, value.var = "pct")
colnames(long16) <- c("fips", "st", "total_votes", "DonaldTrump", "HillaryClinton")
long16$diff <- long16$DonaldTrump - long16$HillaryClinton

#Added an extra leading 0 for FIPS codes that were 4 digits long (enables the merge later)
for (i in seq(nrow(long16))) {
  if (nchar(long16$fips[i])==4) long16$fips[i] <- paste("0",long16$fips[i], sep="")
}

long16$TrumpWin <- ifelse(long16$diff >0, long16$TrumpWin <- 1, long16$TrumpWin <- 0)
long16$TrumpWin <- as.factor(long16$TrumpWin)

####Cleaning FIPS database####
fips.labels$fips <- paste(fips.labels$V2,fips.labels$V3, sep = "")
fips.labels$V2 <- NULL
fips.labels$V3 <- NULL
fips.labels$V5 <- NULL
colnames(fips.labels) <- c("State", "County", "FIPS")

####Merge FIPS database with 2016 election results####
long16 <- merge(long16, fips.labels[-1], by.x = "fips", by.y = "FIPS", all.x = TRUE)

####2012 General Election Data####
#Data import
all_counties_2012 <- read.csv("~/Second Year/DS 4559 - Data Science/Final Project 2/election2016/Data/2012data/all_counties_2012.csv",
                              stringsAsFactors = FALSE)
all_counties_2012 <- subset(all_counties_2012, fips!="fips")
all_counties_2012$votes <- as.numeric(all_counties_2012$votes)
#case insensitive search for rows containing obama or romney
all_counties_2012_romney <- all_counties_2012[grepl('romney',all_counties_2012$candidate,ignore.case=TRUE), ] 
all_counties_2012_obama <- all_counties_2012[grepl('obama',all_counties_2012$candidate,ignore.case=TRUE), ] 

#all_counties_2012_romney <- aggregate (. ~ fips, data=all_counties_2012_romney, FUN=sum)

all_counties_2012 <- rbind(all_counties_2012_romney,all_counties_2012_obama)
all_counties_2012 <- subset(all_counties_2012, fips !="")
summary(all_counties_2012)
summary(all_counties_2012$fips)

#include only first letter
all_counties_2012[,3] <- substring(all_counties_2012[,3], 1, 1) 

all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="o", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="O", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="B", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="b", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="m", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="M", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="r", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="R", "romney")
all_counties_2012 <- aggregate (votes ~ fips+candidate+county, data=all_counties_2012, FUN=sum)

summary(all_counties_2012$candidate) 

all_counties_2012 <- data.frame(lapply(all_counties_2012, as.character), stringsAsFactors=FALSE)
all_counties_2012$votes <- as.numeric(all_counties_2012$votes)

long12 <- dcast(all_counties_2012, fips ~ candidate, value.var = "votes")

for (i in seq(nrow(long12))){
  if (long12$fips[i] == "46113") long12$fips[i] <- "46102"
}

long12$diff <- long12$obama - long12$romney

long12$ObamaWin <- ifelse(long12$diff >0, long12$ObamaWin <- 1, long12$ObamaWin <- 0)
long12$ObamaWin <- as.factor(long12$ObamaWin)

#Note that the percentages calculated do not take into account third-party votes (marginal)
long12$ObamaPer <- long12$obama/(long12$obama + long12$romney)
long12$RomneyPer <- long12$romney/(long12$obama + long12$romney)
long12$diffper <- long12$ObamaPer - long12$RomneyPer

long12 <- merge(long12, fips.labels, by.x = "fips", by.y = "FIPS", all.x = TRUE)

####Lower 48 State US Geographic Data####
us.counties <- counties(c("AL", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME",
                          "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", 
                          "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"), cb = TRUE)

us.counties2 <- fortify(us.counties, region = "GEOID")

us.states <- states(cb = TRUE)
us.states2 <- fortify(us.states, region = "GEOID")
us.states3 <- us.states2[which(us.states2$lat >= 24.396308 & us.states2$lat <= 49.384358 & us.states2$long >= -124.848974 & us.states2$long <= -66.885444),]
```

### Gradient Map
```{r fig.width=16, fig.height = 8, echo = FALSE}
df_merged.us <- merge(us.counties2, long16, by.x = "id", by.y = "fips", all.x = TRUE)

us.ggmap <- ggplot() +
  geom_polygon(data = df_merged.us, aes(x = long, y = lat, group = group, fill = diff), color = "dark grey", size = 0.25) + 
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_gradient(low = "blue", high = "red") + labs(fill = "Trump Margin of Victory")+
  ggtitle("2016 Electoral Map by County") + coord_map("polyconic") + theme_void() 
us.ggmap
```

The gradient within the map represents the margin of victory, from 60% or greater margin of victory for Trump in red to 60% or more margin of victory for Clinton in blue.

Naturally, Clinton has a visible edge in more urban, coastal areas while Trump has an advantage in more rural areas.

***

### Pure Win-Loss
We can clarify the win by county by removing the gradient.

```{r fig.width=16, fig.height = 8, echo = FALSE}
us.ggmap2 <- ggplot() +
  geom_polygon(data = df_merged.us, aes(x = long, y = lat, group = group, fill = TrumpWin), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("blue","red"), labels=c("Clinton", "Trump"),name="County Winner") + 
  ggtitle("2016 Electoral Map by County") + coord_map("polyconic") + theme_void() 
us.ggmap2
```

***

### Interactive
Let's check out a few interactive maps created with `leaflet`. Clinton narrowly won Virginia while Trump narrowly won Pennsylvania. Clicking on the county will give you its name, Trump's margin of victory, as well as each candidate's respective number of votes (and share of the vote).

#### Virginia
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
counties <- counties("VA", cb = TRUE)

#Merge map file with election results based on FIPS code
df_merged <- geo_join(counties, long16, "GEOID", "fips")

#Determine color gradient for the map
pal <- colorNumeric(
  palette = c("blue", "red"),
  domain = df_merged$percent
)

#Determine the content of the pop-up on mouse click
popup <- paste0("<b>", paste(df_merged$County, df_merged$st, sep = ", "), "</b> <br>",
                #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                "<b>Trump Differential: </b>", percent(round(df_merged$diff,2)),
                "<br>", "<b>Trump: </b>", percent(round(df_merged$DonaldTrump,2)), " (",trimws(format(round(df_merged$DonaldTrump*df_merged$total_votes, 0), big.mark = ",")), ")",
                "<br>", "<b>Clinton: </b>", percent(round(df_merged$HillaryClinton,2)), " (",trimws(format(round(df_merged$HillaryClinton*df_merged$total_votes, 0), big.mark = ",")), ")")

```

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged, 
              fillColor = ~pal(diff), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup) %>%
  addLegend(pal = pal, 
            values = df_merged$diff, 
            position = "bottomright", 
            title = "Donald Trump's Advantage",
            labFormat = labelFormat(suffix = "%", transform = function(x) 100 * x))
```


#### Pennsylvania
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
counties2 <- counties("PA", cb = TRUE)
df_merged2 <- geo_join(counties2, long16, "GEOID", "fips")

pal <- colorNumeric(
  palette = c("blue", "red"),
  domain = df_merged2$percent
)

popup.pa <- paste0("<b>", paste(df_merged2$County, df_merged2$st, sep = ", "), "</b> <br>",
                   #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                   "<b>Trump Differential: </b>", percent(round(df_merged2$diff,2)),
                   "<br>", "<b>Trump: </b>", percent(round(df_merged2$DonaldTrump,2)), " (",trimws(format(round(df_merged2$DonaldTrump*df_merged2$total_votes, 0), big.mark = ",")), ")",
                   "<br>", "<b>Clinton: </b>", percent(round(df_merged2$HillaryClinton,2)), " (",trimws(format(round(df_merged2$HillaryClinton*df_merged2$total_votes, 0), big.mark = ",")), ")")

```

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged2, 
              fillColor = ~pal(diff), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup.pa) %>%
  addLegend(pal = pal, 
            values = df_merged2$diff, 
            position = "bottomright", 
            title = "Donald Trump's Advantage",
            labFormat = labelFormat(suffix = "%", transform = function(x) 100 * x))
```


#### United States
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
df_merged.us2 <- geo_join(us.counties, long16, "GEOID", "fips")

pal <- colorNumeric(
  palette = c("blue", "red"),
  domain = df_merged.us2$percent
)

popup.us <- paste0("<b>", paste(df_merged.us2$County, df_merged.us2$st, sep = ", "), "</b> <br>",
                   #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                   "<b>Trump Differential: </b>", percent(round(df_merged.us2$diff,2)),
                   "<br>", "<b>Trump: </b>", percent(round(df_merged.us2$DonaldTrump,2)), " (",trimws(format(round(df_merged.us2$DonaldTrump*df_merged.us2$total_votes, 0), big.mark = ",")), ")",
                   "<br>", "<b>Clinton: </b>", percent(round(df_merged.us2$HillaryClinton,2)), " (",trimws(format(round(df_merged.us2$HillaryClinton*df_merged.us2$total_votes, 0), big.mark = ",")), ")")

```

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged.us2, 
              fillColor = ~pal(diff), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup.us) %>%
  addPolygons(data = us.states, 
              color = "#ffffff",
              fillOpacity = 0, 
              weight = 1, 
              smoothFactor = 0.2, options = pathOptions(clickable = FALSE)) %>%
  addLegend(pal = pal, 
            values = df_merged.us2$diff, 
            position = "bottomright", 
            title = "Donald Trump's Advantage",
            labFormat = labelFormat(suffix = "%", transform = function(x) 100 * x)) %>%
  fitBounds(-124.848974, 24.396308, -66.885444, 49.384358)
```
