---
title: "Exploring the 2016 Election"
author: "Nathan Lin, Andrew Ton, and Mansoor Syed"
date: "December 2016"
output:
  html_document:
    df_print: paged
    
abstract: |
  This data project seeks to document the rise of Donald Trump through the demographics and geography of the United States. We employ techniques in data visualization and machine learning in attempt to characterize, understand, and explain Donald Trump's ascendance to the presidency. Our analysis found that Donald Trump's success in flipping counties previously won by Obama, particularly in pivotal areas such as the Northeast and Midwest, greatly contributed to his victory. Additionially, results from our predictive models indicate that identity politics was more prevalent than ever in the 2016 cycle, with models achieving around 94% accuracy in predicting county-level results based upon US Census data alone.
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
library(reshape2)
library(tigris)
library(ggplot2)
library(leaflet)
library(maps)
library(scales)
library(rgdal)
library(rgeos)
library(sqldf)

####General Notes####
# Use the section-picker at the bottom of the script window to jump between code sections
# Our analysis only considered the lower 48 states and the District of Columbia (excluded Alaska, Hawaii and overseas territories)
# The tigris data requires the below global option if you are running Windows 10
options(tigris_use_cache = FALSE)

####2016 General Election Data####
setwd("~/Second Year/DS 4559 - Data Science/Final Project 2/election2016/Data")

#Read the 2016 general election data + FIPS code database
gen16 <- read.csv("pres16results.csv", header = TRUE, stringsAsFactors = FALSE)
fips.labels <- read.csv("fips_database.csv", header = FALSE, stringsAsFactors = FALSE, colClasses = "character")

#Cleaning general election data
gen16 <- gen16[which(gen16$cand == "Donald Trump" | gen16$cand == "Hillary Clinton"),]
gen16 <- gen16[-c(1,2),]

sapply(gen16,class)

#This county wasn't labeled in the data, so we manually assigned it the county name
for (i in seq(nrow(gen16))){
  if (gen16$fips[i] == "46102") gen16$county[i] <- "Oglala Lakota County"
}

#Creating a separate table with just the state-wide results
gen16.states <- gen16[is.na(gen16$county),]
gen16.states <- gen16.states[-c(103,104),] #removing state FIPS for Alaska
gen16.states$county <- NULL
gen16.states$fips <- NULL

gen16 <- gen16[-is.na(gen16$county)]

#Removing the statewide results from the county data
for (i in seq(nrow(gen16))) {
  if (nchar(gen16$fips[i])<=2) gen16$fips[i] <- NA
}

gen16 <- gen16[-which(is.na(gen16$fips)),]
gen16$pct_report <- NULL
gen16.2 <- gen16

gen16.2$votes <- NULL
gen16.2$lead <- NULL

#Used the reshape2 package to convert the dataframe from long to wide format
long16 <- dcast(gen16.2, fips + st + total_votes ~ cand, value.var = "pct")
colnames(long16) <- c("fips", "st", "total_votes", "DonaldTrump", "HillaryClinton")
long16$diff <- long16$DonaldTrump - long16$HillaryClinton

#Added an extra leading 0 for FIPS codes that were 4 digits long (enables the merge later)
for (i in seq(nrow(long16))) {
  if (nchar(long16$fips[i])==4) long16$fips[i] <- paste("0",long16$fips[i], sep="")
}

long16$TrumpWin <- ifelse(long16$diff >0, long16$TrumpWin <- 1, long16$TrumpWin <- 0)
long16$TrumpWin <- as.factor(long16$TrumpWin)

####Cleaning FIPS database####
fips.labels$fips <- paste(fips.labels$V2,fips.labels$V3, sep = "")
fips.labels$V2 <- NULL
fips.labels$V3 <- NULL
fips.labels$V5 <- NULL
colnames(fips.labels) <- c("State", "County", "FIPS")

####Merge FIPS database with 2016 election results####
long16 <- merge(long16, fips.labels[-1], by.x = "fips", by.y = "FIPS", all.x = TRUE)

####2012 General Election Data####
#Data import
all_counties_2012 <- read.csv("~/Second Year/DS 4559 - Data Science/Final Project 2/election2016/Data/2012data/all_counties_2012.csv",
                              stringsAsFactors = FALSE)
all_counties_2012 <- subset(all_counties_2012, fips!="fips")
all_counties_2012$votes <- as.numeric(all_counties_2012$votes)
#case insensitive search for rows containing obama or romney
all_counties_2012_romney <- all_counties_2012[grepl('romney',all_counties_2012$candidate,ignore.case=TRUE), ] 
all_counties_2012_obama <- all_counties_2012[grepl('obama',all_counties_2012$candidate,ignore.case=TRUE), ] 

#all_counties_2012_romney <- aggregate (. ~ fips, data=all_counties_2012_romney, FUN=sum)

all_counties_2012 <- rbind(all_counties_2012_romney,all_counties_2012_obama)
all_counties_2012 <- subset(all_counties_2012, fips !="")
summary(all_counties_2012)
summary(all_counties_2012$fips)

#include only first letter
all_counties_2012[,3] <- substring(all_counties_2012[,3], 1, 1) 

all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="o", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="O", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="B", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="b", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="m", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="M", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="r", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="R", "romney")
all_counties_2012 <- aggregate (votes ~ fips+candidate+county, data=all_counties_2012, FUN=sum)

summary(all_counties_2012$candidate) 

all_counties_2012 <- data.frame(lapply(all_counties_2012, as.character), stringsAsFactors=FALSE)
all_counties_2012$votes <- as.numeric(all_counties_2012$votes)

long12 <- dcast(all_counties_2012, fips ~ candidate, value.var = "votes")

for (i in seq(nrow(long12))){
  if (long12$fips[i] == "46113") long12$fips[i] <- "46102"
}

long12$diff <- long12$obama - long12$romney

long12$ObamaWin <- ifelse(long12$diff >0, long12$ObamaWin <- 1, long12$ObamaWin <- 0)
long12$ObamaWin <- as.factor(long12$ObamaWin)

#Note that the percentages calculated do not take into account third-party votes (marginal)
long12$ObamaPer <- long12$obama/(long12$obama + long12$romney)
long12$RomneyPer <- long12$romney/(long12$obama + long12$romney)
long12$diffper <- long12$ObamaPer - long12$RomneyPer

long12 <- merge(long12, fips.labels, by.x = "fips", by.y = "FIPS", all.x = TRUE)

####Lower 48 State US Geographic Data####
us.counties <- counties(c("AL", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME",
                          "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", 
                          "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"), cb = TRUE)

us.counties2 <- fortify(us.counties, region = "GEOID")

us.states <- states(cb = TRUE)
us.states2 <- fortify(us.states, region = "GEOID")
us.states3 <- us.states2[which(us.states2$lat >= 24.396308 & us.states2$lat <= 49.384358 & us.states2$long >= -124.848974 & us.states2$long <= -66.885444),]
```

```{r fig.width=16, fig.height = 8, echo = FALSE}
df_merged.us <- merge(us.counties2, long16, by.x = "id", by.y = "fips", all.x = TRUE)

us.ggmap2 <- ggplot() +
  geom_polygon(data = df_merged.us, aes(x = long, y = lat, group = group, fill = TrumpWin), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("blue","red"), labels=c("Clinton", "Trump"),name="County Winner") + 
  coord_map("polyconic") + theme_void() 
us.ggmap2
```

## Motivation

The 2016 General Election for President of the United States was our very first general election in which we were able to vote. Many pundits, politicians, and citizens alike found the contest between Hillary Clinton and Donald Trump to be one of the most bitter presidential races in recent memory, and many more were surprised by the outcome.

Most pollsters and statisticians had Hillary Clinton maintaining a comfortable margin before the election. Yet, Donald Trump emerged victorious as the 45th President of the United States. Our goal is to try to explain why.

We have incorporated US Census demographic data, as well as employing comparisons between Donald Trump's 2016 victory over Hillary Clinton with Barack Obama's 2012 re-election victory against Mitt Romney.


## Content
This website primarily displays the visualizations generated from our analysis coupled with machine learning models. They were created via `leaflet` and `ggplot2` in `R`. Our 2012 election data was obtained from the Huffington Post, and our 2016 election data was sourced from Kaggle.

Code chunks for the graphics and models are not output. Please consult our Github (https://github.com/theNathanLin/election2016) for more information, including our original graphics and code.

Furthermore, our analysis excludes Alaska and Hawaii as well as third-party candidates.


## Navigation
* [2016 Election](election16.html "2016 Election")
    + Static and interactive win/loss maps for the 2016 general election covering Virginia, Pennsylvania, and the United States
* [2012 Election](election12.html "2012 Election")
    + Static maps for the 2012 general election covering the United States
* [2012, 2016 Comparison](election1216.html "2012/2016 Comparisons")
    + Static maps for the 2012 and 2016 general elections covering the United States
    + Comparison of flipped counties from 2012 to 2016
    + Interactive comparisons of Obama and Clinton in the Northeast and Midwest
    + Voter turnout analysis and comparison
* [Hypothetical Scenarios](hypotheticals.html "Hypothetical Scenarios")
    + Interactive comparisons of Obama and Trump in the Northeast and Midwest
    + Static map of an Obama/Trump matchup covering the United States
* [Machine Learning and Predictive Models](models.html "Machine Learning and Predictive Models")
    + Variable importance measures from US Census data
    + Conditional Inference Trees (`ctree`)
    + Sequential Covering/Rule Based Methods (`jRip`)
    + Lazy learning, k-nearest-neighbor (`kNN`)
    + Static maps for 2016 county-by-county predictions covering the United States