---
title: "Exploring the 2016 Election"
author: "Nathan Lin, Andrew Ton, and Mansoor Syed"
date: "December 16, 2016"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    
abstract: |
  What went wrong in 2016? This data project seeks to document the rise of Donald Trump through the demographics and geography of the United States. We employ techniques in data visualization and machine learning in attempt to characterize, understand, and explain Donald Trump's ascendance to the presidency.
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Motivation

The 2016 General Election for President of the United States was our very first general election in which we were able to vote. Many pundits, politicians, and citizens alike found the contest between Hillary Clinton and Donald Trump to be one of the most bitter presidential races in recent memory, and many more were surprised by the outcome.

Most pollsters and statisticians had Hillary Clinton maintaining a comfortable margin before the election. Yet, Donald Trump emerged victorious as the 45th President of the United States. Our goal is to try to explain why.

We have incorporated US Census demographic data, as well as employing comparisons between Donald Trump's 2016 victory with Barack Obama's 2012 re-election against Mitt Romney.

### What You Will Find Below
Below are primarily the visualizations generated from our analysis. They were created via `leaflet` and `ggplot2`. Our 2012 election data was obtained from the Huffington Post, and our 2016 election data was sourced from Kaggle.

Code for the graphics is not included in chunks below. Please consult our Github for more information.

Our analysis excludes Alaska and Hawaii.

***

## 2016 Election
### Margin of Victory

Let's take a look at the 2016 electoral map, divided by county.
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
library(reshape2)
library(tigris)
library(ggplot2)
library(leaflet)
library(maps)
library(scales)
library(rgdal)
library(rgeos)
library(sqldf)

####General Notes####
# Use the section-picker at the bottom of the script window to jump between code sections
# Our analysis only considered the lower 48 states and the District of Columbia (excluded Alaska, Hawaii and overseas territories)
# The tigris data requires the below global option if you are running Windows 10
options(tigris_use_cache = FALSE)

####2016 General Election Data####
setwd("~/Second Year/DS 4559 - Data Science/Final Project/election2016/Data")

#Read the 2016 general election data + FIPS code database
gen16 <- read.csv("pres16results.csv", header = TRUE, stringsAsFactors = FALSE)
fips.labels <- read.csv("fips_database.csv", header = FALSE, stringsAsFactors = FALSE, colClasses = "character")

#Cleaning general election data
gen16 <- gen16[which(gen16$cand == "Donald Trump" | gen16$cand == "Hillary Clinton"),]
gen16 <- gen16[-c(1,2),]

sapply(gen16,class)

#This county wasn't labeled in the data, so we manually assigned it the county name
for (i in seq(nrow(gen16))){
  if (gen16$fips[i] == "46102") gen16$county[i] <- "Oglala Lakota County"
}

#Creating a separate table with just the state-wide results
gen16.states <- gen16[is.na(gen16$county),]
gen16.states <- gen16.states[-c(103,104),] #removing state FIPS for Alaska
gen16.states$county <- NULL
gen16.states$fips <- NULL

gen16 <- gen16[-is.na(gen16$county)]

#Removing the statewide results from the county data
for (i in seq(nrow(gen16))) {
  if (nchar(gen16$fips[i])<=2) gen16$fips[i] <- NA
}

gen16 <- gen16[-which(is.na(gen16$fips)),]
gen16$pct_report <- NULL
gen16.2 <- gen16

gen16.2$votes <- NULL
gen16.2$lead <- NULL

#Used the reshape2 package to convert the dataframe from long to wide format
long16 <- dcast(gen16.2, fips + st + total_votes ~ cand, value.var = "pct")
colnames(long16) <- c("fips", "st", "total_votes", "DonaldTrump", "HillaryClinton")
long16$diff <- long16$DonaldTrump - long16$HillaryClinton

#Added an extra leading 0 for FIPS codes that were 4 digits long (enables the merge later)
for (i in seq(nrow(long16))) {
  if (nchar(long16$fips[i])==4) long16$fips[i] <- paste("0",long16$fips[i], sep="")
}

long16$TrumpWin <- ifelse(long16$diff >0, long16$TrumpWin <- 1, long16$TrumpWin <- 0)
long16$TrumpWin <- as.factor(long16$TrumpWin)

####Cleaning FIPS database####
fips.labels$fips <- paste(fips.labels$V2,fips.labels$V3, sep = "")
fips.labels$V2 <- NULL
fips.labels$V3 <- NULL
fips.labels$V5 <- NULL
colnames(fips.labels) <- c("State", "County", "FIPS")

####Merge FIPS database with 2016 election results####
long16 <- merge(long16, fips.labels[-1], by.x = "fips", by.y = "FIPS", all.x = TRUE)

####2012 General Election Data####
#Data import
all_counties_2012 <- read.csv("~/Second Year/DS 4559 - Data Science/Final Project/election2016/Data/2012data/all_counties_2012.csv",
                              stringsAsFactors = FALSE)
all_counties_2012 <- subset(all_counties_2012, fips!="fips")
all_counties_2012$votes <- as.numeric(all_counties_2012$votes)
#case insensitive search for rows containing obama or romney
all_counties_2012_romney <- all_counties_2012[grepl('romney',all_counties_2012$candidate,ignore.case=TRUE), ] 
all_counties_2012_obama <- all_counties_2012[grepl('obama',all_counties_2012$candidate,ignore.case=TRUE), ] 

#all_counties_2012_romney <- aggregate (. ~ fips, data=all_counties_2012_romney, FUN=sum)

all_counties_2012 <- rbind(all_counties_2012_romney,all_counties_2012_obama)
all_counties_2012 <- subset(all_counties_2012, fips !="")
summary(all_counties_2012)
summary(all_counties_2012$fips)

#include only first letter
all_counties_2012[,3] <- substring(all_counties_2012[,3], 1, 1) 

all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="o", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="O", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="B", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="b", "obama")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="m", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="M", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="r", "romney")
all_counties_2012$candidate <- replace(all_counties_2012$candidate, all_counties_2012$candidate=="R", "romney")
all_counties_2012 <- aggregate (votes ~ fips+candidate+county, data=all_counties_2012, FUN=sum)

summary(all_counties_2012$candidate) 

all_counties_2012 <- data.frame(lapply(all_counties_2012, as.character), stringsAsFactors=FALSE)
all_counties_2012$votes <- as.numeric(all_counties_2012$votes)

long12 <- dcast(all_counties_2012, fips ~ candidate, value.var = "votes")

for (i in seq(nrow(long12))){
  if (long12$fips[i] == "46113") long12$fips[i] <- "46102"
}

long12$diff <- long12$obama - long12$romney

long12$ObamaWin <- ifelse(long12$diff >0, long12$ObamaWin <- 1, long12$ObamaWin <- 0)
long12$ObamaWin <- as.factor(long12$ObamaWin)

#Note that the percentages calculated do not take into account third-party votes (marginal)
long12$ObamaPer <- long12$obama/(long12$obama + long12$romney)
long12$RomneyPer <- long12$romney/(long12$obama + long12$romney)
long12$diffper <- long12$ObamaPer - long12$RomneyPer

long12 <- merge(long12, fips.labels, by.x = "fips", by.y = "FIPS", all.x = TRUE)

####Lower 48 State US Geographic Data####
us.counties <- counties(c("AL", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME",
                          "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", 
                          "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"), cb = TRUE)

us.counties2 <- fortify(us.counties, region = "GEOID")

us.states <- states(cb = TRUE)
us.states2 <- fortify(us.states, region = "GEOID")
us.states3 <- us.states2[which(us.states2$lat >= 24.396308 & us.states2$lat <= 49.384358 & us.states2$long >= -124.848974 & us.states2$long <= -66.885444),]
```

### Gradient Map
```{r fig.width=20, fig.height = 10, echo = FALSE}
df_merged.us <- merge(us.counties2, long16, by.x = "id", by.y = "fips", all.x = TRUE)

us.ggmap <- ggplot() +
  geom_polygon(data = df_merged.us, aes(x = long, y = lat, group = group, fill = diff), color = "dark grey", size = 0.25) + 
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_gradient(low = "blue", high = "red") + labs(fill = "Trump Margin of Victory")+
  ggtitle("2016 Electoral Map by County") + coord_map("polyconic") + theme_void() 
us.ggmap
```

The gradient within the map represents the margin of victory, from 60% or greater margin of victory for Trump in red to 60% or more margin of victory for Clinton in blue.

Naturally, Clinton has a visible edge in more urban, coastal areas while Trump has an advantage in more rural areas.

***

### Pure Win-Loss
We can clarify the win by county by removing the gradient.

```{r fig.width=20, fig.height = 10, echo = FALSE}
us.ggmap2 <- ggplot() +
  geom_polygon(data = df_merged.us, aes(x = long, y = lat, group = group, fill = TrumpWin), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("blue","red"), labels=c("Clinton", "Trump"),name="County Winner") + 
  ggtitle("2016 Electoral Map by County") + coord_map("polyconic") + theme_void() 
us.ggmap2
```

***

### Interactive
Let's check out a few interactive maps created with `leaflet`.

#### Virginia
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
counties <- counties("VA", cb = TRUE)

#Merge map file with election results based on FIPS code
df_merged <- geo_join(counties, long16, "GEOID", "fips")

#Determine color gradient for the map
pal <- colorNumeric(
  palette = c("blue", "red"),
  domain = df_merged$percent
)

#Determine the content of the pop-up on mouse click
popup <- paste0("<b>", paste(df_merged$County, df_merged$st, sep = ", "), "</b> <br>",
                #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                "<b>Trump Differential: </b>", percent(round(df_merged$diff,2)),
                "<br>", "<b>Trump: </b>", percent(round(df_merged$DonaldTrump,2)), " (",trimws(format(round(df_merged$DonaldTrump*df_merged$total_votes, 0), big.mark = ",")), ")",
                "<br>", "<b>Clinton: </b>", percent(round(df_merged$HillaryClinton,2)), " (",trimws(format(round(df_merged$HillaryClinton*df_merged$total_votes, 0), big.mark = ",")), ")")

```

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged, 
              fillColor = ~pal(diff), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup) %>%
  addLegend(pal = pal, 
            values = df_merged$diff, 
            position = "bottomright", 
            title = "Donald Trump's Advantage",
            labFormat = labelFormat(suffix = "%", transform = function(x) 100 * x))
```


#### Pennsylvania
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
counties2 <- counties("PA", cb = TRUE)
df_merged2 <- geo_join(counties2, long16, "GEOID", "fips")

pal <- colorNumeric(
  palette = c("blue", "red"),
  domain = df_merged2$percent
)

popup.pa <- paste0("<b>", paste(df_merged2$County, df_merged2$st, sep = ", "), "</b> <br>",
                   #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                   "<b>Trump Differential: </b>", percent(round(df_merged2$diff,2)),
                   "<br>", "<b>Trump: </b>", percent(round(df_merged2$DonaldTrump,2)), " (",trimws(format(round(df_merged2$DonaldTrump*df_merged2$total_votes, 0), big.mark = ",")), ")",
                   "<br>", "<b>Clinton: </b>", percent(round(df_merged2$HillaryClinton,2)), " (",trimws(format(round(df_merged2$HillaryClinton*df_merged2$total_votes, 0), big.mark = ",")), ")")

```

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged2, 
              fillColor = ~pal(diff), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup.pa) %>%
  addLegend(pal = pal, 
            values = df_merged2$diff, 
            position = "bottomright", 
            title = "Donald Trump's Advantage",
            labFormat = labelFormat(suffix = "%", transform = function(x) 100 * x))
```


#### United States
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
df_merged.us2 <- geo_join(us.counties, long16, "GEOID", "fips")

pal <- colorNumeric(
  palette = c("blue", "red"),
  domain = df_merged.us2$percent
)

popup.us <- paste0("<b>", paste(df_merged.us2$County, df_merged.us2$st, sep = ", "), "</b> <br>",
                   #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                   "<b>Trump Differential: </b>", percent(round(df_merged.us2$diff,2)),
                   "<br>", "<b>Trump: </b>", percent(round(df_merged.us2$DonaldTrump,2)), " (",trimws(format(round(df_merged.us2$DonaldTrump*df_merged.us2$total_votes, 0), big.mark = ",")), ")",
                   "<br>", "<b>Clinton: </b>", percent(round(df_merged.us2$HillaryClinton,2)), " (",trimws(format(round(df_merged.us2$HillaryClinton*df_merged.us2$total_votes, 0), big.mark = ",")), ")")

```

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged.us2, 
              fillColor = ~pal(diff), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup.us) %>%
  addLegend(pal = pal, 
            values = df_merged.us2$diff, 
            position = "bottomright", 
            title = "Donald Trump's Advantage",
            labFormat = labelFormat(suffix = "%", transform = function(x) 100 * x))
```

***

## 2012 Election
### Margin of Victory
```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
df_merged.us12 <- merge(us.counties2, long12, by.x = "id", by.y = "fips", all.x = TRUE)

####2012-2016 Differences####
#Counties Obama Won and Clinton Lost
diff.1216 <- merge(long12[1:5], long16[c(1,2,3,4,5,7,8)], by = "fips")
diff.1216$trump <- diff.1216$total_votes*diff.1216$DonaldTrump
diff.1216$clinton <- diff.1216$total_votes*diff.1216$HillaryClinton
diff.1216[c(4,7:9)] <- NULL
diff.1216 <- diff.1216[c(1,5, 7,2,3,4,6,8,9)]

diff.1216$flip <- ifelse(diff.1216$ObamaWin == 1 & diff.1216$TrumpWin==1, 2, 
                         ifelse(diff.1216$ObamaWin == 0 & diff.1216$TrumpWin==0, 3, 
                                ifelse(diff.1216$TrumpWin==1, 1, 0)))
diff.1216$flip <- as.factor(diff.1216$flip)

diff.1216$obama_trump <- ifelse(diff.1216$obama > diff.1216$trump, 1, 0)
diff.1216$obama_trump <- as.factor(diff.1216$obama_trump)

df_merged.us2.12 <- merge(us.counties2, diff.1216[c(1,2,3,10,11)], by.x = "id", by.y = "fips", all.x = TRUE)
```


```{r fig.width=20, fig.height = 10, echo = FALSE}
us.ggmap12 <- ggplot() +
  geom_polygon(data = df_merged.us12, aes(x = long, y = lat, group = group, fill = diffper), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_gradient(low = "red", high = "blue") + labs(fill = "Obama Margin of Victory")+
  ggtitle("2012 Electoral Map by County") + coord_map("polyconic") + theme_void() 
us.ggmap12
```

***

###Pure Win-Loss
Like the procedure above, we can remove the gradient and look at the winner by county.

```{r fig.width=20, fig.height = 10, echo = FALSE}
us.ggmap2.12 <- ggplot() +
  geom_polygon(data = df_merged.us12, aes(x = long, y = lat, group = group, fill = ObamaWin), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("red","blue"), labels=c("Romney", "Obama"),name="County Winner") + 
  ggtitle("2012 Electoral Map by County") + coord_map("polyconic") + theme_void()
us.ggmap2.12
```

It is certainly quite evident that Obama was able to win a significantly greater number of counties in 2012 compared to Clinton in 2016. We will run through some exact comparisons below.

***

## 2012 and 2016 Comparisons
### Map Comparisons
Let's compare the win-loss maps by county from 2012 to 2016.

##### 2012
```{r fig.width=20, fig.height = 10, echo = FALSE}
us.ggmap2.12 <- ggplot() +
  geom_polygon(data = df_merged.us12, aes(x = long, y = lat, group = group, fill = ObamaWin), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("red","blue"), labels=c("Romney", "Obama"),name="County Winner") + 
  ggtitle("2012 Electoral Map by County") + coord_map("polyconic") + theme_void()
us.ggmap2.12
```


##### 2016
```{r fig.width=20, fig.height = 10, echo = FALSE}
us.ggmap2 <- ggplot() +
  geom_polygon(data = df_merged.us, aes(x = long, y = lat, group = group, fill = TrumpWin), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("blue","red"), labels=c("Clinton", "Trump"),name="County Winner") + 
  ggtitle("2016 Electoral Map by County") + coord_map("polyconic") + theme_void() 
us.ggmap2
```

***

### 2016 Flipped Counties
How successful were each of the candidates in flipping counties from their previous victor? Below is a graph of the 2016 electoral map, with flips highlighted in yellow for Trump (Trump winning a county won by Obama in 2012) and green for Clinton (Clinton winning a county won by Romney in 2012).

```{r fig.width=20, fig.height = 10, echo = FALSE}
us.ggmap.1216 <- ggplot() +
  geom_polygon(data = df_merged.us2.12, aes(x = long, y = lat, group = group, fill = flip), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("blue","red", "yellow", "green"), 
                    labels=c("Clinton","Trump", "Trump Flip", "Clinton Flip"),name="County Winner") + 
  ggtitle("2012 Electoral Map by County, Flips") + coord_map("polyconic") + theme_void()
us.ggmap.1216
```

We can clean up the colors by marking the unchanged counties in grey and re-marking the flipped counties for clarity.
```{r fig.width=20, fig.height = 10, echo = FALSE}
us.ggmap.1216.2 <- ggplot() +
  geom_polygon(data = df_merged.us2.12, aes(x = long, y = lat, group = group, fill = flip), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("gray","gray", "red", "blue"), 
                    labels=c("Clinton","Trump", "Trump Flip", "Clinton Flip"),name="County Winner") + 
  ggtitle("2012 Electoral Map by County, Flips") + coord_map("polyconic") + theme_void()
us.ggmap.1216.2
```

#### Table of Flipped Counties by State
We can further drill down and take a look at the geographic breakdown of flipped counties by state.

##### Trump
```{r, echo = FALSE, rows.print = 38, warning = FALSE, message=FALSE}
flipped.counties.trump <- sqldf("select st, count(st) from 'diff.1216' where flip == 2 group by st")
colnames(flipped.counties.trump) <- c("State", "Counties Flipped")
flipped.counties.trump <- flipped.counties.trump[order(flipped.counties.trump$`Counties Flipped`, decreasing = TRUE),]
flipped.counties.trump <- rbind(flipped.counties.trump, c("Total", sum(flipped.counties.trump$'Counties Flipped')))

flipped.counties.trump
```


##### Clinton
```{r, echo = FALSE, rows.print = 12, warning = FALSE, message=FALSE}
flipped.counties.clinton <- sqldf("select st, count(st) from 'diff.1216' where flip == 3 group by st")
colnames(flipped.counties.clinton) <- c("State", "Counties Flipped")
flipped.counties.clinton <- flipped.counties.clinton[order(flipped.counties.clinton$`Counties Flipped`, decreasing = TRUE),]
flipped.counties.clinton <- rbind(flipped.counties.clinton, c("Total", sum(flipped.counties.clinton$'Counties Flipped')))

flipped.counties.clinton
```

As evidenced from the tables above, Trump was much more successful than Clinton in flipping counties from 2012 to 2016.

***

### Interactive
Given the information from above, it looks as if Clinton lost significant numbers of counties in the Northeast and Midwest that were once held by Obama. Let's explore in greater detail

#### Northeast
```{r echo=FALSE}
co.diff <- diff.1216
co.diff$co.diff <- co.diff$obama - co.diff$clinton
co.diff$co.diff2 <- co.diff$obama - co.diff$trump

counties.ME.NY <- counties(state = c("ME","NY"), cb = TRUE)

df_merged.ME.NY <- geo_join(counties.ME.NY, co.diff, "GEOID", "fips")

pal.ME.NY <- colorBin("Blues", df_merged.ME.NY$co.diff, 8, pretty = FALSE)

popup.ME.NY <- paste0("<b>", paste(df_merged.ME.NY$County, df_merged.ME.NY$st, sep = ", "), "</b> <br>",
                   #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                   "<b>Obama Differential: </b>", format(round(df_merged.ME.NY$co.diff,0), big.mark = ","),
                   "<br>", "<b>Obama: </b>",trimws(format(round(df_merged.ME.NY$obama, 0), big.mark = ",")),
                   "<br>", "<b>Clinton: </b>", trimws(format(round(df_merged.ME.NY$clinton, 0), big.mark = ",")),
                   "<br>", "<b>Trump: </b>", trimws(format(round(df_merged.ME.NY$trump, 0), big.mark = ",")))

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged.ME.NY, 
              fillColor = ~pal.ME.NY(co.diff), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup.ME.NY) %>%
  addLegend(pal = pal.ME.NY, 
            values = df_merged.ME.NY$co.diff, 
            position = "bottomright", 
            title = "Obama-Clinton Vote Differential (+ favors Obama)")
```


#### Midwest
```{r echo=FALSE}
counties.midwest <- counties(state = c("MN","MI", "WI", "IA", "IL"), cb = TRUE)
df_merged.midwest <- geo_join(counties.midwest, co.diff, "GEOID", "fips")

pal.midwest <- colorBin("Blues", df_merged.midwest$co.diff, 8, pretty = FALSE)

popup.midwest <- paste0("<b>", paste(df_merged.midwest$County, df_merged.midwest$st, sep = ", "), "</b> <br>",
                     #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                     "<b>Obama Differential: </b>", format(round(df_merged.midwest$co.diff,0), big.mark = ","),
                     "<br>", "<b>Obama: </b>",trimws(format(round(df_merged.midwest$obama, 0), big.mark = ",")),
                     "<br>", "<b>Clinton: </b>", trimws(format(round(df_merged.midwest$clinton, 0), big.mark = ",")),
                     "<br>", "<b>Trump: </b>", trimws(format(round(df_merged.midwest$trump, 0), big.mark = ",")))

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged.midwest, 
              fillColor = ~pal.midwest(co.diff), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup.midwest) %>%
  addLegend(pal = pal.midwest, 
            values = df_merged.midwest$co.diff, 
            position = "bottomright", 
            title = "Obama-Clinton Vote Differential (+ favors Obama)")
```

***

## Hypotheticals
Even though we understand that the circumstances surrounding both the 2012 and 2016 elections were different, how well could Obama have performed in the 2016 election had he gone against Trump and maintained the same performance that he did in 2012?

### Interactive
#### Northeast
```{r echo=FALSE}
pal.ME.NY2 <- colorBin("Reds", df_merged.ME.NY$co.diff2, 8, pretty = FALSE)

popup.ME.NY2 <- paste0("<b>", paste(df_merged.ME.NY$County, df_merged.ME.NY$st, sep = ", "), "</b> <br>",
                      #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                      "<b>Obama Differential: </b>", format(round(df_merged.ME.NY$co.diff2,0), big.mark = ","),
                      "<br>", "<b>Obama: </b>",trimws(format(round(df_merged.ME.NY$obama, 0), big.mark = ",")),
                      "<br>", "<b>Trump: </b>", trimws(format(round(df_merged.ME.NY$trump, 0), big.mark = ",")),
                      "<br>", "<b>Clinton: </b>", trimws(format(round(df_merged.ME.NY$clinton, 0), big.mark = ",")))

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged.ME.NY, 
              fillColor = ~pal.ME.NY2(co.diff2), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup.ME.NY2) %>%
  addLegend(pal = pal.ME.NY2, 
            values = df_merged.ME.NY$co.diff2, 
            position = "bottomright", 
            title = "Obama-Trump Vote Differential (+ favors Obama)")
```

#### Midwest
```{r echo=FALSE}
pal.midwest2 <- colorBin("Reds", df_merged.midwest$co.diff2, 8, pretty = FALSE)

popup.midwest2 <- paste0("<b>", paste(df_merged.midwest$County, df_merged.midwest$st, sep = ", "), "</b> <br>",
                       #"<b>FIPS Code: </b>", df_merged$GEOID, "<br>", 
                       "<b>Obama Differential: </b>", format(round(df_merged.midwest$co.diff2,0), big.mark = ","),
                       "<br>", "<b>Obama: </b>",trimws(format(round(df_merged.midwest$obama, 0), big.mark = ",")),
                       "<br>", "<b>Trump: </b>", trimws(format(round(df_merged.midwest$trump, 0), big.mark = ",")),
                       "<br>", "<b>Clinton: </b>", trimws(format(round(df_merged.midwest$clinton, 0), big.mark = ",")))

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = df_merged.midwest, 
              fillColor = ~pal.midwest2(co.diff2), 
              color = "#b2aeae",
              fillOpacity = 0.8, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = popup.midwest2) %>%
  addLegend(pal = pal.ME.NY2, 
            values = df_merged.midwest$co.diff2, 
            position = "bottomright", 
            title = "Obama-Trump Vote Differential (+ favors Obama)")
```

### Head-to-Head Hypothetical Matchup
Here's what the US county electoral map would have looked like if we matched Obama's 2012 performance to Trump's 2016 victory.

```{r fig.width=20, fig.height = 10, echo=FALSE}
us.ggmap.obamatrump <- ggplot() +
  geom_polygon(data = df_merged.us2.12, aes(x = long, y = lat, group = group, fill = obama_trump), color = "dark grey", size = 0.25) +
  geom_path(data = us.states3, aes(x=long, y=lat, group =group), color = "white", size = 1) +
  scale_fill_manual(values = c("red","blue"), 
                    labels=c("Trump","Obama"),name="County Winner") + 
  ggtitle("Obama-Trump Hypothetical Matchup") + coord_map("polyconic") + theme_void() 
us.ggmap.obamatrump
```